Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-07-11 manuscript.tex
addressing reviewer 1, comment 3
addressing reviewer 1, comment 4

When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions. Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree. The improper use of resolution parameter may lead to two types of errors: grouping of cells that are in fact distinct and creating an inconsistent hierarchy.

@@

The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme, not limited to the solution given by the model itself. Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data. We sought to exploit this property to address the task of annotating cells according to a reference sample. To this end we analysed datasets from \cite{mereu_2020}, which includes mixtures of human PBMC and HEK293T cells profiled with various technologies. We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartz-seq2 or MARS-seq. These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task.

@@

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c c ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{nSBM} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:06 & 00:13 & 00:36 \\ 
 \hline
 Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:10 & 00:19 & 00:45 \\
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:20 & 00:34 & 02:14 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:23 & 00:40 & 03:02 \\
 \hline
 Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:14 & 00:26 & 01:07 \\ 
% \hline
% Islets \cite{Segerstolpe_2016} & 2108 & 32023 & 00:20 & 00:42 & 03:13 \\
 \hline
 Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:37 & 01:27 & 05:52\\ 
 \hline
 Mouse Cortex \cite{Zeisel_2015} & 3005 & 54460 & 00:59 & 00:53 & 07:32 \\
 \hline
 Endocrinogenesis \cite{Bastidas_Ponce_2019} & 3696 & 74670 & 01:15 & 01:29 & 10:56 \\
 \hline
 Baron Pancreas \cite{Baron_2016} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
 \hline
 Airzani Liver \cite{Aizarani_2019} & 10368 & 354440 & 04:13 & 09:47 & 1:42:23 \\
 \hline
 Planaria \cite{plass_2018} & 21612 & 173667 & 05:41 & 13:52 & 1:20:40 \\
 \hline
\end{tabular}
\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. All approaches fit 100 models. Number of nodes and edges refer to the structure of the \emph{k}NN graph as built by \texttt{scanpy}. Times are expressed in MM:SS. }
\label{Table_runtime}
\end{table}

>>>>

When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, \emph{i.e.} communities are consistently grouped at lower resolutions. Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree. The improper use of resolution parameter may lead to two types of errors: grouping of cells that are in fact distinct and creating an inconsistent hierarchy.

@@

The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme, not limited to the solution given by the model itself. Differences in entropy can be used to perform model selection, hence we can choose which model better describes the data. We sought to exploit this property to address the task of annotating cells according to a reference sample. To this end we analysed datasets from \cite{mereu_2020}, which includes mixtures of human PBMC and HEK293T cells profiled with various technologies. We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartz-seq2 or MARS-seq. These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task.

@@

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c | c c c ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{nSBM} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:06 & 00:13 & 00:36 \\ 
 \hline
 Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:10 & 00:19 & 00:45 \\
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:20 & 00:34 & 02:14 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:23 & 00:40 & 03:02 \\
 \hline
 Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:14 & 00:26 & 01:07 \\ 
% \hline
% Islets \cite{Segerstolpe_2016} & 2108 & 32023 & 00:20 & 00:42 & 03:13 \\
 \hline
 Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:37 & 01:27 & 05:52\\ 
 \hline
 Mouse Cortex \cite{Zeisel_2015} & 3005 & 54460 & 00:59 & 00:53 & 07:32 \\
 \hline
 Endocrinogenesis \cite{Bastidas_Ponce_2019} & 3696 & 74670 & 01:15 & 01:29 & 10:56 \\
 \hline
 Baron Pancreas \cite{Baron_2016} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
 \hline
 Airzani Liver \cite{Aizarani_2019} & 10368 & 354440 & 04:13 & 09:47 & 1:42:23 \\
 \hline
 Planaria \cite{plass_2018} & 21612 & 173667 & 05:41 & 13:52 & 1:20:40 \\
 \hline
\end{tabular}
\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. All approaches fit 100 models. Number of nodes and edges refer to the structure of the \emph{k}NN graph as built by \texttt{scanpy}. Times are expressed in MM:SS. }
\label{Table_runtime}
\end{table}
